Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-04-25 paper/main.tex
asi nejake male zmeny (netusim jake :-))
polishing of sections 1--3

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%\documentclass[a4paper,UKenglish,cleveref, autoref, anonymous, thm-restate]{lipics-v2021}
%
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

\newcommand{\Nset}{{\mathbb{N}_0}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
%\def\Inf{\ensuremath{\mathsf{Inf}}}
%\def\Fin{\ensuremath{\mathsf{Fin}}}
\DeclareMathOperator{\Inf}{\mathsf{Inf}}
\DeclareMathOperator{\Fin}{\mathsf{Fin}}
\newcommand{\minf}{\mathit{inf}}
\newcommand{\mks}{\mathit{mks}}
%\newcommand{\oM}{\widehat{M}}
\newcommand{\oM}{N}
\newcommand{\rem}{\mathit{nm}}
\newcommand{\mcycle}{\mathit{cycle}}
\newcommand{\msat}{\mathit{satisfies}}
\newcommand{\Te}{T_{\vec{e}}}
% \newcommand{\pre}{{\leftarrow\!\!S}}
% \newcommand{\post}{{S\!\!\rightarrow}}
\newcommand{\pre}{{S_\mathit{pre}}}
\newcommand{\post}{{S_\mathit{post}}}
\def\false{\mathit{false}}
\def\true{\mathit{true}}
\newcommand{\mA}{\mathcal{A}}
\newcommand{\mB}{\mathcal{B}}

@@

\begin{abstract}
  \emph{Transition-based Emerson-Lei automata (TELA)} become a popular
  formalism that generalizes many traditional kinds of automata over
  infinite words including Büchi, co-Büchi, Rabin, Streett, and parity
  automata. Transitions in a TELA are labelled with acceptance marks
  and its accepting formula is a positive Boolean combination of terms
  saying that a particular mark has to be visited infinitely or
  finitely often. % in an accepting run.
  Algorithms processing these automata are often very sensitive to the
  number of %different
  acceptance marks. We introduce a new technique reducing the number
  of %different
  acceptance marks in TELA with use of \emph{quantified Boolean
    formulas (QBF)}.
  % : if a constructed QBF formula is satisfiable then
  % its model encodes an equivalent accepting formula with less
  % acceptance marks.
  In fact, we present three formula constructions trading speed for
  higher reduction efficiency. We evaluated our reduction technique on
  TELA produced by state-of-the-art tools of the libraries Owl and
  Spot and by the tool \texttt{ltl3tela}. The technique reduced some
  acceptance marks in automata produced by all the tools. The achieved
  reduction on automata produced by tools of the Owl library was
  between 30 and 40\%.
\end{abstract}

@@

In 1987, Emerson and Lei~\cite{emerson.87.scp} introduced automata
over infinite words where acceptance conditions are arbitrary
combinations of acceptance primitives saying that a certain set of
states should be visited finitely often or infinitely often. In 2015,
the same kind of acceptance condition was described in the \emph{Hanoi
  omega-automata format (HOAF)}~\cite{babiak.15.cav}. The only
difference is that the acceptance primitives talk about finitely or
infinitely often visited acceptance marks rather than sets of states.
Acceptance marks are placed on transitions and each mark identifies
the set of transitions containing this mark. Hence, these automata are
called \emph{transition-based Emerson-Lei automata (TELA)} and they
generalize many traditional kinds of automata over infinite words
including Büchi, co-Büchi, Rabin, Streett, and parity automata.

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with simpler acceptance
conditions. Further, algorithms handling TELA can automatically handle
all automata with traditional acceptance conditions. TELA can
be obtained for example by translating LTL formulas by
\texttt{ltl2dela} (known as Delag)~\cite{muller.17.gandalf}),
\texttt{ltl3tela}~\cite{major.19.atva},
\texttt{ltl2dgra} (known as Rabinizer~4)~\cite{kretinsky.18.cav},
or \texttt{ltl2tgba} which is a part of the Spot library~\cite{duret.16.atva2}.
There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity
automata~\cite{renkin.20.atva,casares.22.tacas}.

@@

We have implemented our reduction technique in a tool called
\telatko. We show that the tool can reduce acceptance marks in
automata produced by Delag~\cite{muller.17.gandalf},
Rabinizer~4~\cite{kretinsky.18.cav} (both included in the Owl library),
Spot~\cite{duret.16.atva2}
and \texttt{ltl3tela}~\cite{major.19.atva}.  While the reduction is relatively
modest for TELA produced by \texttt{ltl3tela} and Spot,
it is substantial for automata produced by tools of the Owl library.

@@

A run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $\minf(\pi)$ denote the set of transitions that appear
infinitely many times in $\pi$. Run $\pi$ is \emph{accepting} iff
$\minf(\pi)$ satisfies the formula $\varphi$, where a set of $T$ of
transitions satisfies $\Inf m$ iff $m\in\mks(T)$ and it satisfies
$\Fin m$ iff $m\not\in\mks(T)$. The \emph{language} of automaton $\mA$ is
the set
$L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of
}\mA\textrm{ over }u\}$. Two automata $\mA,\mB$ are \emph{equivalent}
if $L(\mA)=L(\mB)$.
% such that $\mA$ has
% an accepting run over $u$.

An acceptance formula $\varphi$ is in \emph{disjunctive normal formal
  (DNF)} if it is a disjunction of clauses, where each clause is a
conjunction of some terms of the form $\Fin m$ or $\Inf m$. Formula
$\false$ corresponds to the disjunction of zero clauses and formula
$\true$ corresponds to one clause with zero terms. Each acceptance
formula can be transformed into an equivalent formula in DNF.

A \emph{path} from a state $p$ to a state $q$ is a finite sequence of
adjacent transitions of the form
$\rho=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots(q_{n-1},u_{n-1},M_{n-1},q_n)\in\delta^+$
such that $p=q_0$ and $q=q_n$. A nonempty set of states $S\subseteq Q$
is called a nontrivial \emph{strongly connected component (SCC)} if
for each $p,q\in S$ there is a path from $p$ to $q$. An SCC S is
\emph{maximal} if there is no SCC $S'$ satisfying $S\subsetneq S'$. In
the rest of this paper, SCC always refers to a maximal SCC. Given a
set of states $S\subseteq Q$, $\delta_S=\delta\cap (S\times \Sigma \times 2^M \times S)$
denotes the set of all transitions between states in $S$. Further, for
each mark $m\in M$, let $\delta_m$ denotes the set of all transitions
marked with $m$, i.e., $\delta_m=\{t\in\delta\mid m\in
\mks(t)\}$.
% \todo{The sets $\delta_m$ are called \emph{acceptance sets}. Given
% an SCC $S$ and a mark $m\in M$, by $\delta_{S,m}$ we denote the set
% $\delta_{S,m}=\delta_S\cap\delta_m$.}
A set of transitions $T\subseteq\delta$ is called a \emph{cycle} is
there exists a path from a state $p$ to the same state containing all
transitions of $T$ (potentially multiple times) and no transition
outside $T$. Finally, we assume that each TELA $\mA$ contains only
states $q$ that are reachable from the initial state $q_I$ (i.e.,
$q=q_I$ or there is a path from $q_I$ to $q$) as states that are not
reachable from $q_I$ can be eliminated without any impact on $L(\mA)$.

@@

\emph{Quantified Boolean formulas (QBF)} are Boolean formulas
extended with universal and existential quantification over
propositional variables. We assume that subformulas of the form
$\forall x.\psi$ and $\exists x.\psi$ do not contain another
quantification of variable $x$ inside $\psi$. The semantics of
$\forall x.\psi$ and $\exists x.\psi$ is given by equivalences
\[
  \begin{array}{rcl}
    \forall x.\psi &~\equiv~& \psi[x\rightarrow\true]~\wedge~\psi[x\rightarrow\false]\\
    \exists x.\psi &~\equiv~& \psi[x\rightarrow\true]~\vee~   \psi[x\rightarrow\false]
  \end{array}
\]
where $\psi[x\rightarrow \rho]$ denotes the formula $\psi$ with all
occurrences of $x$ simultaneously replaced by $\rho$. The equivalences
imply that each QBF can be transformed into an equivalent Boolean
formula. However, the size of this Boolean formula can be exponential
in the size of the original QBF. Let $V$ be the set of all
propositional variables. A mapping $\mu:V\rightarrow\{\true,\false\}$
is a \emph{model} of a QBF formula $\varphi$ iff it is a satisfying
assignment of the equivalent Boolean formula. A QBF formula is
\emph{satisfiable} iff it has a model.

@@

We start with several simple observations. Acceptance of a run $\pi$
is fully determined by the set $\minf(\pi)$.
% of acceptance marks that
% appear infinitely often on $\pi$, which is in turn fully determined by
% the set $\minf(\pi)$ of transitions appearing infinitely often in
% $\pi$.
Further, each set $\minf(\pi)$ for a run $\pi$ is a cycle and vice
versa. Hence, our reduction methods is looking for a new acceptance
formula $\psi$ and function $\rem$ such that for each cycle
$T\subseteq\delta$, it holds that $T$ satisfies $\varphi$ if and only
if $\rem(T)$ satisifes $\psi$. This can be roughly denoted by the
formula
\[
  \forall T\subseteq\delta~.~\mcycle(T)\implies\big(\msat_\varphi(T) ~\iff~ \msat_{\psi,\rem}(T)\big).
\]
In fact, this corresponds to the shape of the QBF formulas we will
construct. As we are looking for $\psi$ and $\rem$ such that the
formula holds, the subformula $\msat_{\psi,\rem}(T)$ contains many
free variables representing possible instances of $\psi$ and
$\rem$. If the formula is satisfiable, then its models encode desired
instances of $\psi$ and $\rem$.  In the following, we assume that we
are looking for a new acceptance formula $\psi$ in DNF. Note that our
reduction method can be easily adapted to look for $\psi$ of a
different shape.

Now we describe the construction of QBF formulas in detail. The
construction is parametrized by two integers $C,K\ge 0$, where $K$ is
the desired number of acceptance marks and $C$ is the number of
clauses of $\psi$. For simplicity, we assume that the reduced
automaton will use the acceptance marks $N_K=\{1,2,\ldots,K\}$.
%Further, we consider only subsets $T\subseteq\bigcup_{\textrm{SCC}~S}\delta_S$
First we describe Boolean variables used in the constructed QBF
formulas.
\begin{itemize}
\item For each transition $t\in\delta$, variable $e_t$ says whether
  $t$ is in the current set $T$ or not.
  \[e_t=\left\{
      \begin{array}{ll}
        1 & \textrm{if }t\in T\\
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
%  In fact, we consider variables 
\item For each transition $t\in\delta$ and acceptance mark $k\in N_K$,
  variable $n_{t,k}$ says whether $k$ is on the transition
  $\rem(t)$ or not.
  \[n_{t,k}=\left\{
      \begin{array}{ll}
        1 & \textrm{if }k\in\mks(\rem(t))\\
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
\item For each $c\in\{1,2,\ldots,C\}$ and acceptance mark $k\in N_K$,
  variables $i_{c,k}$ and $f_{c,k}$ say whether the $c^\textrm{th}$
  clause of $\psi$ contains terms $\Inf k$ or $\Fin k$, respectively.
  \[i_{c,k}=\left\{
      \begin{array}{ll}
        1 & \textrm{if the $c^\textrm{th}$ clause of $\psi$ contains $\Inf k$}\\
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
  \[f_{c,k}=\left\{
      \begin{array}{ll}
        1 & \textrm{if the $c^\textrm{th}$ clause of $\psi$ contains $\Fin k$}\\
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]

@@

The subformula $\msat_\varphi(\vec{e})$ says whether $T$ satisfies the
original acceptance formula $\varphi$ and it is derived directly from
$\varphi$. Recall that $T$ satisfies $\Inf m$ iff $m\in\mks(T)$, which
means that $T$ contains some transition with mark $m$. As the
transitions with mark $m$ form the set $\delta_m$, $\Inf m$ can be
expresses by $\bigvee_{t\in\delta_m}e_t$. Similarly, $T$ satisfies
$\Fin m$ iff $m\not\in\mks(T)$, which can be expresses by
$\bigwedge_{t\in\delta_m}\neg e_t$. Hence, $\msat_\varphi(\vec{e})$
arises from $\varphi$ by replacing
\begin{itemize}
\item all terms of the form $\Inf m$ by $\bigvee_{t\in\delta_m}e_t$ and
\item all terms of the form $\Fin m$ by $\bigwedge_{t\in\delta_m}\neg e_t$.
\end{itemize}

@@

It remains to define the subformula $\mcycle(\vec{e})$. Let $\Te$
denote the set of transitions represented by $\vec{e}$. The original
intended meaning of $\mcycle(\vec{e})$ is
\[
  \mcycle(\vec{e}) \iff \Te\textrm{ is a cycle}.
\]
In fact, only the direction ``$\Longleftarrow$'' is required for the
correctness of our reduction method. If $\Te$ is not a cycle and
$\mcycle(\vec{e})$ holds, then the required equivalence
$\msat_\varphi(\vec{e}) \iff
\msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})$ on this valuation of
$\vec{e}$ adds a superfluous constraint on $\psi$ represented by
$\vec{i},\vec{f}$ and $\rem$ represented by $\vec{n}$. These
superfluous constraints can lead to loss of reduction opportunities,
but not to incorrectness. This observation allows us to trade the
precision of $\mcycle(\vec{e})$ for simplicity of the formula.

In the following, we define three versions of $\mcycle(\vec{e})$:
\begin{itemize}
\item $\mcycle_1(\vec{e})$ is a lightweight version, which only says
  that $\Te$ is nonempty. It does not use the information about
  automaton structure, but it comes with an interesting simplification
  of the whole formula $\Phi_{C,K}$.
\item $\mcycle_2(\vec{e})$ is an intermediate version. It says that $\Te$
  is nonempty and every transition in $\Te$ has to have a preceding
  and a succeeding transition in $\Te$, which is a necessary condition
  for being a cycle, but not a sufficient one.
\item $\mcycle_3(\vec{e})$ is a strict version saying that $\Te$ is a
  cycle. Unfortunately, it uses additional universally quantified
  variables corresponding to automata states. Transformation of
  $\Phi_{C,K}$ to prenex normal form turns the quantifiers to
  existential ones and the resulting formula thus contains quantifier
  alternation.
\end{itemize}
We write $\Phi_{j,C,K}$ when we want to emphasize that the particular
formula $\Phi_{C,K}$ contains the version $\mcycle_j(\vec{e})$.

@@

Formally, we define an equivalence $\sim\subseteq\delta\times\delta$
on transitions such that $t_1\sim t_2$ whenever $\mks(t_1)=\mks(t_2)$.
\begin{lemma}
  Assume that there is a function $\rem$ and a formula $\psi$ such that 
  \[
    \textrm{for every nonempty set $T\subseteq\delta$ it holds }
      (T~\msat~\varphi~\iff~\rem(T)~\msat~\psi).\tag{$1$}\label{eq:lem1} 
  \]  
  Then there exists a function $\rem'$ that respects % the equivalence
  $\sim$ (i.e.,
  if $t_1\sim t_2$ then $\rem'(t_1)=\rem'(t_2)$) and
  \[
    \textrm{for every nonempty set $T\subseteq\delta$ it holds }
    (T~\msat~\varphi~\iff~\rem'(T)~\msat~\psi).\tag{$2$}\label{eq:lem2} 
  \]  
\end{lemma}
\begin{proof}
  Let $\rem$ be a function and $\psi$ a formula such that
  \eqref{eq:lem1} holds. To construct the function $\rem'$, we first
  select one transition from each equivalence class of $\sim$. For
  every transition $t\in\delta$, by $\overline{t}$ we denote the
  selected transition equivalent to $t$. We define the function
  $\rem'$ as $\rem'(t)=\rem(\overline{t})$. Clearly, $\rem'$ respects
  the equivalence $\sim$. It remains to show that \eqref{eq:lem2}
  holds for $\rem'$ and $\psi$.

>>>>

\documentclass[a4paper,UKenglish,cleveref,autoref,thm-restate]{lipics-v2021}
%\documentclass[a4paper,UKenglish,cleveref, autoref, anonymous, thm-restate]{lipics-v2021}
%
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

\newcommand{\Nset}{{\mathbb{N}_0}}
\newcommand{\todo}[1]{\marginpar{\textcolor{red}{#1}}}
%\def\Inf{\ensuremath{\mathsf{Inf}}}
%\def\Fin{\ensuremath{\mathsf{Fin}}}
\DeclareMathOperator{\Inf}{\mathsf{Inf}}
\DeclareMathOperator{\Fin}{\mathsf{Fin}}
\newcommand{\minf}{\mathit{inf}}
\newcommand{\mks}{\mathit{mks}}
%\newcommand{\oM}{\widehat{M}}
\newcommand{\oM}{N}
\newcommand{\rem}{\mathit{nm}}
\newcommand{\mcycle}{\mathit{cycle}}
\newcommand{\msat}{\mathit{satisfies}}
\newcommand{\Te}{T_{\vec{e}}}
% \newcommand{\pre}{{\leftarrow\!\!S}}
% \newcommand{\post}{{S\!\!\rightarrow}}
\newcommand{\pre}{{S_\mathit{pre}}}
\newcommand{\post}{{S_\mathit{post}}}
\def\false{\mathit{false}}
\def\true{\mathit{true}}
\newcommand{\mA}{\mathcal{A}}
\newcommand{\mB}{\mathcal{B}}

@@

\begin{abstract}
  \emph{Transition-based Emerson-Lei automata (TELA)} became a popular
  formalism that generalizes many traditional kinds of automata over
  infinite words including Büchi, co-Büchi, Rabin, Streett, and parity
  automata. Transitions in a TELA are labelled with acceptance marks
  and its accepting formula is a positive Boolean combination of terms
  saying that a particular mark has to be visited infinitely or
  finitely often. % in an accepting run.
  Algorithms processing these automata are often very sensitive to the
  number of %different
  acceptance marks. We introduce a new technique reducing the number
  of %different
  acceptance marks in TELA with use of \emph{quantified Boolean
    formulas (QBF)}.
  % : if a constructed QBF formula is satisfiable then
  % its model encodes an equivalent accepting formula with less
  % acceptance marks.
  In fact, we present three formula constructions
  % trading speed for higher reduction efficiency.
  with increasing reduction potential and formula complexity, and thus
  also decreasing speed of their satisfiability solving. We evaluated
  our reduction technique on TELA produced by state-of-the-art tools
  of the libraries Owl and Spot and by the tool \texttt{ltl3tela}. The
  technique reduced some acceptance marks in automata produced by all
  the tools. On automata with more than one acceptance mark produced
  by tools Delag and Rabinizer~4 of the Owl library, our technique reduced
  reduced 27.5\% and 38.4\% of acceptance marks, respectively. \todo{zkontrolovat!}
  % The achieved
  % reduction of acceptance marks on automata produced by tools of the Owl library was
  % between 25 and 40\%.
\end{abstract}

@@

Automata over infinite words like Büchi, Rabin, Streett, or parity
automata play a crucial role in many algorithms related to the
concurrency theory. In particular, they are used in specification,
verification, analysis, monitoring, and synthesis of various systems
with infinite behaviour. In 1987, Emerson and
Lei~\cite{emerson.87.scp} introduced automata over infinite words
where acceptance conditions are arbitrary combinations of acceptance
primitives saying that a certain set of states should be visited
finitely often or infinitely often. In 2015, the same kind of
acceptance condition was described in the \emph{Hanoi omega-automata
  format (HOAF)}~\cite{babiak.15.cav}. The only difference is that the
acceptance primitives talk about finitely or infinitely often visited
acceptance marks rather than sets of states.  Acceptance marks are
placed on transitions and each mark identifies the set of transitions
containing this mark. Hence, these automata are called
\emph{transition-based Emerson-Lei automata (TELA)} and they
generalize many traditional kinds of automata over infinite words
including Büchi, co-Büchi, Rabin, Streett, and parity automata.

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with simpler acceptance
conditions. Further, algorithms handling TELA can automatically handle
all automata with traditional acceptance conditions. TELA can be
obtained for example by translating LTL formulas by \texttt{ltl2dela}
(known as Delag)~\cite{muller.17.gandalf},
\texttt{ltl3tela}~\cite{major.19.atva}, \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}, or \texttt{ltl2tgba} which is a
part of the Spot library~\cite{duret.16.atva2}.  There are also
algorithms processing these automata, for example the emptiness
check~\cite{baier.19.atva} or translation of TELA to parity
automata~\cite{renkin.20.atva,casares.22.tacas}.

@@

We have implemented our reduction technique in a tool called
\telatko. We show that the tool can reduce acceptance marks in
automata produced by Delag~\cite{muller.17.gandalf},
Rabinizer~4~\cite{kretinsky.18.cav} (both included in the Owl library),
Spot~\cite{duret.16.atva2}
and \texttt{ltl3tela}~\cite{major.19.atva}.  While the reduction is relatively
modest on TELA produced by \texttt{ltl3tela} and Spot,
it is substantial on automata produced by the tools of the Owl library.

@@

A \emph{run} $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\ldots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\ldots\in\delta^{\omega}$ where
$q_0=q_I$. Let $\minf(\pi)$ denote the set of transitions that appear
infinitely many times in $\pi$. Run $\pi$ is \emph{accepting} iff
$\minf(\pi)$ satisfies the formula $\varphi$, where a set $T$ of
transitions satisfies $\Inf m$ iff $m\in\mks(T)$ and it satisfies
$\Fin m$ iff $m\not\in\mks(T)$. The \emph{language} of automaton $\mA$ is
the set
$L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of
}\mA\textrm{ over }u\}$. Two automata $\mA,\mB$ are \emph{equivalent}
if $L(\mA)=L(\mB)$.
% such that $\mA$ has
% an accepting run over $u$.

An acceptance formula $\varphi$ is in \emph{disjunctive normal formal
  (DNF)} if it is a disjunction of clauses, where each clause is a
conjunction of terms of the form $\Fin m$ or $\Inf m$.  Each
acceptance formula can be transformed into an equivalent formula
in~DNF. Formula $\false$ corresponds to the disjunction of zero
clauses and formula $\true$ corresponds to one clause with zero terms.

A \emph{path} from a state $p$ to a state $q$ is a finite sequence of
adjacent transitions of the form
$\rho=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\ldots(q_{n-1},u_{n-1},M_{n-1},q_n)\in\delta^+$
such that $p=q_0$ and $q=q_n$. A nonempty set of states $S\subseteq Q$
is called a nontrivial \emph{strongly connected component (SCC)} if
for each $p,q\in S$ there is a path from $p$ to $q$. An SCC S is
\emph{maximal} if there is no SCC $S'$ satisfying $S\subsetneq S'$. In
the rest of this paper, SCC always refers to a maximal SCC. Given a
set of states $S\subseteq Q$, let $\delta_S=\delta\cap (S\times \Sigma \times 2^M \times S)$
denote the set of all transitions between states in $S$. Further, for
each mark $m\in M$, let $\delta_m=\{t\in\delta\mid m\in
\mks(t)\}$ denote the set of all transitions
marked with $m$.
% \todo{The sets $\delta_m$ are called \emph{acceptance sets}. Given
% an SCC $S$ and a mark $m\in M$, by $\delta_{S,m}$ we denote the set
% $\delta_{S,m}=\delta_S\cap\delta_m$.}
A set of transitions $T\subseteq\delta$ is called a \emph{cycle} is
there exists a path from a state $p$ to the same state containing each
transition of $T$ at least once and no transition outside
$T$. Finally, we assume that each TELA $\mA$ contains only states $q$
that are reachable from the initial state $q_I$ (i.e., $q=q_I$ or
there is a path from $q_I$ to $q$) as states that are not reachable
from $q_I$ can be eliminated without any impact on $L(\mA)$.

@@

\emph{Quantified Boolean formulas (QBF)} are Boolean formulas
extended with universal and existential quantification over
propositional variables. We assume that subformulas of the form
$\forall x.\psi$ and $\exists x.\psi$ do not contain another
quantification of variable $x$ inside $\psi$. The semantics of
$\forall x.\psi$ and $\exists x.\psi$ is given by equivalences
\[
  \begin{array}{rcl}
    \forall x.\psi &~\equiv~& \psi[x\rightarrow\true]~\wedge~\psi[x\rightarrow\false]\\
    \exists x.\psi &~\equiv~& \psi[x\rightarrow\true]~\vee~   \psi[x\rightarrow\false]
  \end{array}
\]
where $\psi[x\rightarrow \rho]$ denotes the formula $\psi$ with all
occurrences of $x$ simultaneously replaced by $\rho$. The equivalences
imply that each QBF can be transformed into an equivalent Boolean
formula. However, the size of this Boolean formula can be exponential
in the size of the original QBF. Let $V$ be the set of all
propositional variables. A mapping $\mu:V\rightarrow\{\true,\false\}$
is a \emph{model} of a QBF formula $\varphi$ iff it is a satisfying
assignment of an equivalent Boolean formula. A QBF formula is
\emph{satisfiable} iff it has a model.

@@

We start with several simple observations. Acceptance of a run $\pi$
is fully determined by the set $\minf(\pi)$.
% of acceptance marks that
% appear infinitely often on $\pi$, which is in turn fully determined by
% the set $\minf(\pi)$ of transitions appearing infinitely often in
% $\pi$.
Further, each set $\minf(\pi)$ for a run $\pi$ is a cycle and vice
versa. Hence, our reduction methods is looking for a new acceptance
formula $\psi$ and function $\rem$ such that for each cycle
$T\subseteq\delta$, it holds that $T$ satisfies $\varphi$ if and only
if $\rem(T)$ satisifes $\psi$. This can be roughly denoted by the
formula
\[
  \forall T\subseteq\delta~.~\mcycle(T)\implies\big(\msat_\varphi(T) ~\iff~ \msat_{\psi,\rem}(T)\big).
\]
In fact, this corresponds to the shape of the QBF formulas we will
construct. As we are looking for $\psi$ and $\rem$ such that the
formula holds, the subformula $\msat_{\psi,\rem}(T)$ contains many
free variables representing possible instances of $\psi$ and
$\rem$. If the formula is satisfiable, then each of its models encodes
a desired instance of $\psi$ and $\rem$.  In the following, we assume
that we are looking for a new acceptance formula $\psi$ in DNF. Note
that our reduction method can be easily adapted to look for $\psi$ of
a different shape.

Now we describe the construction of QBF formulas in detail. The
construction is parametrized by two integers $C,K\ge 0$, where $K$ is
the desired number of acceptance marks and $C$ is the number of
clauses of $\psi$. For simplicity, we assume that the reduced
automaton will use the acceptance marks $N_K=\{1,2,\ldots,K\}$.
%Further, we consider only subsets $T\subseteq\bigcup_{\textrm{SCC}~S}\delta_S$
We start with a description of Boolean variables used in the
constructed QBF formulas.
\begin{itemize}
\item For each transition $t\in\delta$, variable $e_t$ says whether
  $t$ is in the current set $T$ or not.
  \[e_t=\left\{
      \begin{array}{ll}
        1 & \textrm{if }t\in T\\
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
%  In fact, we consider variables 
\item For each transition $t\in\delta$ and acceptance mark $k\in N_K$,
  variable $n_{t,k}$ says whether $k$ is on the transition
  $\rem(t)$ or not.
  \[n_{t,k}=\left\{
      \begin{array}{ll}
        1 & \textrm{if }k\in\mks(\rem(t))\\
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
\item For each $c\in\{1,2,\ldots,C\}$ and acceptance mark $k\in N_K$,
  variables $i_{c,k}$ and $f_{c,k}$ say whether the $c^\textrm{th}$
  clause of $\psi$ contains terms $\Inf k$ or $\Fin k$, respectively.
  \[i_{c,k}=\left\{
      \begin{array}{ll}
        1 & \textrm{if the $c^\textrm{th}$ clause of $\psi$ contains $\Inf k$}\\
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
  \[f_{c,k}=\left\{
      \begin{array}{ll}
        1 & \textrm{if the $c^\textrm{th}$ clause of $\psi$ contains $\Fin k$}\\
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]

@@

The subformula $\msat_\varphi(\vec{e})$ says whether $T$ satisfies the
original acceptance formula $\varphi$ and it is derived directly from
$\varphi$. Recall that $T$ satisfies $\Inf m$ iff $m\in\mks(T)$, which
means that $T$ contains some transition with mark $m$. As the
transitions with mark $m$ form the set $\delta_m$, $\Inf m$ can be
expressed by $\bigvee_{t\in\delta_m}e_t$. Similarly, $T$ satisfies
$\Fin m$ iff $m\not\in\mks(T)$, which can be expressed by
$\bigwedge_{t\in\delta_m}\neg e_t$. Hence, $\msat_\varphi(\vec{e})$
arises from $\varphi$ by replacing
\begin{itemize}
\item all terms of the form $\Inf m$ by $\bigvee_{t\in\delta_m}e_t$ and
\item all terms of the form $\Fin m$ by $\bigwedge_{t\in\delta_m}\neg e_t$.
\end{itemize}

@@

It remains to define the subformula $\mcycle(\vec{e})$. Let $\Te$
denote the set of transitions represented by $\vec{e}$. The original
intended meaning of $\mcycle(\vec{e})$ is
\[
  \mcycle(\vec{e}) \iff \Te\textrm{ is a cycle}.
\]
In fact, only the direction ``$\Longleftarrow$'' is needed for the
correctness of our reduction method. If there are some valuations of
$\vec{e}$ such that $\mcycle(\vec{e})$ holds and $\Te$ is not a cycle,
then we will superfluously require the equivalence
$\msat_\varphi(\vec{e}) \iff
\msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})$ on these valuations.
% adds a superfluous constraint on $\psi$ (represented by
% $\vec{i},\vec{f}$) and $\rem$ (represented by $\vec{n}$).
These superfluous constraints can lead to loss of reduction
opportunities, but not to incorrectness. This observation allows us to
trade the precision of $\mcycle(\vec{e})$ for its simplicity.
formula.
%
We define three versions of $\mcycle(\vec{e})$:
\begin{itemize}
\item $\mcycle_1(\vec{e})$ is a lightweight version, which only says
  that $\Te$ is nonempty. It does not use the information about
  automaton structure, but it comes with an interesting simplification
  of the whole formula $\Phi_{C,K}$.
\item $\mcycle_2(\vec{e})$ is an intermediate version. It says that $\Te$
  is nonempty and every transition in $\Te$ has % to have
  a preceding
  and a succeeding transition in $\Te$, which is a necessary condition
  for being a cycle, but not a sufficient one.
\item $\mcycle_3(\vec{e})$ is a strict version saying that $\Te$ is a
  cycle. Unfortunately, it uses additional universally quantified
  variables corresponding to automata states. Transformation of
  $\Phi_{C,K}$ to prenex normal form turns the quantifiers to
  existential ones and the resulting formula thus contains quantifier
  alternation.
\end{itemize}
We write $\Phi_{j,C,K}$ %when we want
to emphasize that a particular
formula $\Phi_{C,K}$ contains the version $\mcycle_j(\vec{e})$.

@@

Formally, we define an equivalence ${\sim}\subseteq\delta\times\delta$
on transitions such that $t_1\sim t_2$ whenever $\mks(t_1)=\mks(t_2)$.
\begin{lemma}
  Assume that there is a function $\rem$ and a formula $\psi$ such that 
  \[
    \textrm{for every nonempty set $T\subseteq\delta$ it holds }
      (T~\msat~\varphi~\iff~\rem(T)~\msat~\psi).\tag{$1$}\label{eq:lem1} 
  \]  
  Then there exists a function $\rem'$ that respects % the equivalence
  $\sim$ (i.e.,
  if $t_1\sim t_2$ then $\rem'(t_1)=\rem'(t_2)$) and
  \[
    \textrm{for every nonempty set $T\subseteq\delta$ it holds }
    (T~\msat~\varphi~\iff~\rem'(T)~\msat~\psi).\tag{$2$}\label{eq:lem2} 
  \]  
\end{lemma}
\begin{proof}
  Let $\rem$ be a function and $\psi$ a formula such that
  \eqref{eq:lem1} holds. To construct the function $\rem'$, we first
  select one transition from each equivalence class of $\sim$. For
  every transition $t\in\delta$, by $\overline{t}$ we denote the
  selected transition equivalent to $t$. We define the function
  $\rem'$ as $\rem'(t)=\rem(\overline{t})$. Clearly, $\rem'$ respects
  the equivalence $\sim$. It remains to show that \eqref{eq:lem2}
  holds for $\rem'$ and $\psi$.
